Versioned Queries over RDF Archives:\\All You Need is SPARQL?

### 2020-09-09 paper.tex
first submitted version

\begin{abstract}
Although many prominent Linked Datasets are in a state of constant evolution, historical data are rarely published in a standard way. Though a number of works have recently addressed the issue of querying and managing archives of versioned RDF data, most propose dedicated solutions, which incurs a high deployment cost. We rather explore solutions for representing archives of versioned RDF data using the SPARQL standard without modification, allowing existing implementations to be used. For querying, we consider two versioning primitives: viewing results from a particular version, or viewing results that change between two consecutive versions. We identify and compare several models by which historical versions of an RDF graph can be stored as named graphs, and discuss how base queries can be rewritten in each model per the two aforementioned primitives. We then present a performance comparison of these models and baseline (non-)versioned queries with respect to 23 weekly versions of Wikidata and 232 user-defined queries, using Virtuoso as a reference SPARQL implementation.
\end{abstract}

@@

A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added. Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change~\cite{KaferAUOH13}. It is in this context that various works have looked at versioning in the context of RDF/SPARQL~\cite{VolkelG06,TappoletB09,Grandi10,GraubeHU14,KhuranaD16}, with recent works proposing \textit{RDF archives}~\cite{FernandezPU15,Cerdeira-PenaFF16,BahriLA18,FernandezUPK19,TaelmanSHMV19} that manage RDF graphs and their historical changes, allowing for querying across different versions of the graph. Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated. While these represent important advances, many such proposals are by their nature incompatible with the existing SPARQL infrastructure. Furthermore, benchmarks have mostly focused on synthetic data. Hence there is still a considerable gap between these research works and putting RDF archives into practice.

@@

In order to bridge this gap, we highlight that -- in theory at least -- there is no need for specialised indexes, query languages, etc., but rather that the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, etc., as well as broad deployment on the Web. For example, SPARQL named graphs can be used to track different versions of individual graphs. However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''. This claim raises the research question: how inefficient will pure SPARQL be in this setting? Though Fernandez at al.~\cite{FernandezUPK19} do discuss how pure SPARQL could be used, to the best of our knowledge, this question has not been empirically explored. If a pure SPARQL solution could be found with relatively little overhead versus base queries (over a single current version) then the benefits would be significant in terms of being able to leverage existing SPARQL infrastructure -- i.e., languages, protocols, implementations -- for managing and querying RDF archives.

In this paper, we present preliminary empirical results addressing this research question. Specifically we take the Virtuoso SPARQL engine~\cite{Erling12} and  evaluate five different representations of RDF archives for XX different weekly versions of truthy Wikidata dumps~\cite{VrandecicK14}. We then implement query rewriting mechanisms for each representation in order to support single-version queries and sequential single-delta queries that retrieve all solutions over an indicated version and all solutions that differ between two sequential versions, respectively. We then perform experiments to compare the sizes of indexes, the costs of indexing new versions, and the costs of query evaluation for base queries over the current version versus rewritten versioned queries.

@@

Various temporal extensions for RDF/SPARQL have been proposed in literature. Gutierrez et al.~\cite{GutierrezHV07} describe a framework that allows time representation in RDF using annotations. Pugliese et al.~\cite{PuglieseUS08} more generally explore time annotations in RDF, proposing semantics, queries and specialised indexes. Zimmerman et al.~\cite{ZimmermannLPS12} provide a general framework for annotating RDF that generalises temporal annotations; they further propose a query language called AnQL for querying annotated RDF. Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc. Related to temporality, a number of systems support versioning for RDF, including SemVersion~\cite{VolkelG06}, POI~\cite{TzitzikasTA08}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.

@@

In summary, the past years have seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data. However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes. Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data. The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc. A number of authors have acknowledged that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}; however, we are not aware of any empirical results along these lines. Our work presents some preliminary results that aim to fill this gap and to better understand the costs and limitations of using vanilla SPARQL to host RDF archives.

@@

We first establish some preliminaries and notation. RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B). In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes in RDF graphs~\cite{ZeginisTC11}. Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$. An RDF graph $G$ is then a set of triples. An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.

A triple pattern $(s,p,o) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V)$ is an RDF triple that permits variables from a set $\V$ to appear in any position. A solution is a partial mapping $\mu : \V \rightarrow \I \cup \L$. We denote by $\dom{\mu}$ the \textit{domain} of $\mu$, i.e., the set of variables for which $\mu$ is defined. We say that two solutions $\mu_1, \mu_2$ are compatible, denoted $\mu_1 \sim \mu_2$, if and only if $\mu_1(v) = \mu_2(v)$ for all $v \in \dom{\mu_1} \cap \dom{\mu_2}$.

A basic graph pattern $B$ is a set of triple patterns. Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$. The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$.

SPARQL queries are based on basic graph patterns and a number of relational operators. Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows. A basic graph pattern $B$ is a graph pattern. Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$, $\sminus{P}{Q}$ and $\ssel{P}{V}$ are graph patterns. We define the semantics of these graph patterns in Table~\ref{tab:sparql}.

\begin{table}[t]
{\scriptsize
\begin{align*}
\sand{P}{Q}(G) \da &\, P(G) \Join Q(G) & M_1 \Join M_2 \da &\, \{ \mu_1 \cup \mu_2 \mid \mu_1 \in M_1,\, \mu_2 \in M_2,\, \com{\mu_1}{\mu_2} \} \\
\suni{P}{Q}(G) \da &\, P(G) \cup Q(G) & M_1 \cup M_2 \da &\, \{ \mu \mid \mu \in M_1\text{ or }\mu \in M_2 \} \\
\sminus{P}{Q}(G) \da &\, P(G) \setminus Q(G) &  M_1 \setminus M_2 \da &\, \{ \mu_1 \in M_1 \mid \nexists \mu_2 \in M_2 : \com{\mu_1}{\mu_2} \} \\
%M_1 \loj M_2 \da\, & (M_1 \bowtie M_2) \cup (M_1 \setminus M_2) \\
%\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
\end{align*}}
\caption{Semantics for SPARQL query operators where $M$, $M_1$ and $M_2$ denote sets of solutions, $V$ denotes a set of variables\label{tab:sparql}}
\end{table}

In this paper, we rely not only on RDF graphs, but also on SPARQL datasets to represent RDF archives. A SPARQL dataset $D \coloneqq \{ G, (n_1,G_1), \ldots (n_m,G_m) \}$ consists of a default (RDF) graph $G$ and a set of named graphs, each being a pair $(n_i,G_i)$ where $n_i \in \I$, $G_i$ is an RDF graph, and $n_i \neq n_j$ for $1 \leq i \leq m$, $1 \leq j \leq m$, $i \neq j$. For simplicity, we will consider a SPARQL dataset $D$ to be a set of quads of the form $(G \times \{*\}) \cup (G_1 \times \{n_1\}) \cup \ldots \cup (G_m \times \{n_m\})$, where $* \notin \I \cup \L \cup \V$ is a special symbol denoting the default graph.\footnote{We thus assume a \textit{quad store}, and disallow empty named graphs.} Different named graphs can be queried using a \texttt{GRAPH} operator, creating quad patterns. A quad pattern $(s,p,o,g) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V) \times (\I \cup \{ *\})$ extends a triple pattern with a fourth element that may be an IRI or *. A basic named graph pattern $C$ is then a set of quad patterns whose evaluation on an SPARQL dataset $D$ is given as $C(D) \coloneqq \{ \mu \mid \mu(C) \subseteq D\text{ and }\dom{\mu} = \vars{C} \}$. Following the SPARQL standard~\cite{sparql11}, we translate a triple pattern $(s,p,o)$ to a quad pattern $(s,p,o,*)$ such that it is evaluated only on the default graph.\footnote{In concrete SPARQL syntax, quad patterns are expressed with the $\texttt{GRAPH}$ keyword.} The semantics of the operators defined in Table~\ref{tab:sparql} then remain unchanged simply allowing $P$ and $Q$ to now represent basic named graph patterns, defined to be graph patterns.

SPARQL provides two operators to initialise a SPARQL dataset: \texttt{FROM} and \texttt{FROM NAMED}. We combine both for brevity into one operator. A graph pattern $P$ is considered to be a query. Likewise if $P$ is a graph pattern, and $M$ and $N$ are sets of IRIs, then $\sfrom{P}{M}{N}$ is also a query. The evaluation of $\sfrom{P}{M}{N}$ on a dataset $D$ is defined as $\sfrom{P}{M}{N}(D) \coloneqq P(D_{M,N})$, where the query dataset $D_{M,N} \coloneqq (\cup_{m \in M} D(m) \times \{ *\}) \cup  (\cup_{n \in N} D(n) \times n)$ is formed from a default graph containing the union\footnote{Since we do not allow blank nodes, a union or RDF merge is equivalent.} of graphs with a name $m \in M$, and all the named graphs in $D$ with a graph named $n \in N$. We will be able to use these operators to define query datasets that capture specific versions in an RDF archive.

@@

\paragraph{Independent Copies (IC):} Perhaps the most intuitive representation is to store each constituent graph of an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that deterministically identify the version. This will result in relatively simple (and thus likely efficient) rewritten queries, but can be expected to occupy a lot of space, particularly in cases where a small proportion of triples change from version to version. We thus consider the $IC$ representation to be a baseline for comparison.

\paragraph{Change-Based (CB):} The core idea of $CB$ representations is to store only triples that change from a given reference version. Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$. We then consider four $CB$ representations based on the following transformations of an RDF archive $\mathcal{G} = (G_1,\ldots,G_n)$:
%
{\footnotesize
\begin{align*}
\mathcal{G}^n_1 \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1) & G_i & = (G_1 \cup \Delta_i^1) \setminus \Delta_1^i & \text{for }1 < i \leq n\\
\mathcal{G}^n_{n-1} \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_{n-1}^n,\Delta_n^{n-1}) & G_i & = (G_{i-1} \cup \Delta_i^{i-1}) \setminus \Delta_{i-1}^i & \text{for }1 < i \leq n \\
\mathcal{G}^1_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_n^1,\Delta_1^n) & G_i & = (G_n \cup \Delta_i^n) \setminus \Delta_n^i & \text{for }1 \leq i < n \\
\mathcal{G}^{n-1}_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_2^1,\Delta_1^2) & G_i & = (G_{i+1} \cup \Delta_i^{i+1}) \setminus \Delta_{i+1}^i & \text{for }1 \leq i < n
\end{align*}%
}%
%
As can be seen for the definitions of $G_i$, these transformations are lossless: we can retrieve any version of the graph from any such transformation. The first two transformations start with the earliest version as a base. The first encodes deltas always with respect to the first version. The second encodes deltas with respect to the previous version. The latter two transformations start with the latest version. The third encodes deltas with respect to the latest version. The second encodes deltas with respect to the subsequent version. Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses''$\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$. Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.

In terms of space we expect $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$ to be the most efficient as they always encode deltas from a neighbouring version. However, in terms of query rewriting, we expect $\mathcal{G}^n_{1}$ and $\mathcal{G}^{1}_n$ to be more efficient as they do not require a recursive construction of all intermediate versions towards the base version. In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas. On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.\footnote{One may note that these four representations are analogous to differential backups, incremental backups, reverse-differential backups, and reverse-incremental backups.}

\paragraph{Timestamp-Based (TB):} The intuition of the $TB$ representation is to associate each triple with the versions in which it is contained. Along these lines, we denote by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.} Let $\mathcal{N} \coloneqq \{ N \mid \exists (s,p,o) \in G_i : \mathcal{G}(s,p,o) = N, 1 \leq i \leq n  \}$ denote the family of sets of versions associated with some triple in $(s,p,o)$. We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$. However, care must be taken as the number of named graphs could potentially reach $2^n$. Though this worst-case is unlikely to happen in practice (it would require triples often being intermittently added and removed across versions), a more practical option is to apply interval encoding and create a named graph for each interval~\cite{FernandezPU15}. More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears. The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of ``intermittent'' triples appearing in multiple named graphs for different intervals. Empty intervals can be omitted.

>>>>

\begin{abstract}
We explore solutions for representing archives of versioned RDF data using the SPARQL standard and off-the-shelf engines. We consider six representations of RDF archives based on named graphs, and describe how input queries can be rewritten to return solutions for a particular version, or solutions that change between versions. We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries, using Virtuoso as the SPARQL engine.
\end{abstract}

@@

A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added. Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change~\cite{KaferAUOH13}. It is in this context that various works have looked at versioning in the context of RDF/SPARQL~\cite{VolkelG06,TappoletB09,Grandi10,GraubeHU14,KhuranaD16}, with recent works proposing \textit{RDF archives}~\cite{FernandezPU15,Cerdeira-PenaFF16,BahriLA18,FernandezUPK19,TaelmanSHMV19} that manage RDF graphs and their historical changes, allowing for querying across different versions of the graph. Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated. While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating an barrier for adoption.

@@

In fact, the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, and deployment. SPARQL named graphs can, for exampe, be used to track different versions of individual graphs. However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''. This raises a question: how inefficient will such queries be? If a pure SPARQL solution could be found with reasonable performance, existing SPARQL infrastructure could be leveraged for RDF archives.

In this paper, we present preliminary empirical results addressing this research question. Specifically we look at six representations of RDF archives using named graphs and propose query rewriting mechanisms for them. We then evaluate and compare these representations for an RDF archive of 8 versions of Wikidata~\cite{VrandecicK14}. Our experiments compare the sizes of the indexes generated, the time taken for indexing, and the relative costs of query evaluation.

@@

%Various temporal extensions for RDF/SPARQL have been proposed in literature. Gutierrez et al.~\cite{GutierrezHV07} describe a framework that allows time representation in RDF using annotations. Pugliese et al.~\cite{PuglieseUS08} more generally explore time annotations in RDF, proposing semantics, queries and specialised indexes. Zimmerman et al.~\cite{ZimmermannLPS12} provide a general framework for annotating RDF that generalises temporal annotations; they further propose a query language called AnQL for querying annotated RDF. Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc. Related to temporality, a number of systems support versioning for RDF, including SemVersion~\cite{VolkelG06}, POI~\cite{TzitzikasTA08}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.

Various temporal extensions for RDF have been proposed in literature based on annotations~\cite{GutierrezHV07,PuglieseUS08,ZimmermannLPS12}. Proposed temporal extensions for SPARQL include $\tau$-SPARQL~\cite{TappoletB09}, T-SPARQL~\cite{Grandi10}, SPARQL-ST~\cite{PerryJS11}, SPARQL\textsuperscript{T}~\cite{ZanioloGACG18}, etc. Related to temporality, a number of systems support versioning for RDF, including SemVersion~\cite{VolkelG06}, POI~\cite{TzitzikasTA08}, x-RDF-3x~\cite{NeumannW10}, R43ples~\cite{GraubeHU14}, Dydra~\cite{AndersonB16} and Ostrich~\cite{TaelmanSV18}.

@@

The past years have thus seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data. However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes, creating an obstacle for adoption. %Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data. The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc. 
A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}. This paper describes preliminary experiments to gain insights into how (in)efficient vanilla SPARQL and off-the-shelf engines are for hosting RDF archives.

@@

We establish some preliminaries and notation. RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B). In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes~\cite{ZeginisTC11}. Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$. An RDF graph $G$ is then a set of triples. An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.

A triple pattern $t \coloneqq (s,p,o) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V)$ is an RDF triple that permits variables from $\V$ to appear in any position. We denote by $\vars{t}$ the variables of $t$. A solution is a partial mapping $\mu : \V \rightarrow \I \cup \L$. We denote by $\dom{\mu}$ the \textit{domain} of $\mu$, i.e., the set of variables for which $\mu$ is defined. We say that two solutions $\mu_1, \mu_2$ are compatible, denoted $\mu_1 \sim \mu_2$, if and only if $\mu_1(v) = \mu_2(v)$ for all $v \in \dom{\mu_1} \cap \dom{\mu_2}$. We denote by $\mu(t)$ the image of $t$ under $\mu$, replacing each variable $v \in \dom{\mu} \cap \vars{t}$ by $\mu(v)$ in $t$. We denote by $t(G) \coloneqq \{ \mu \mid \mu(t) \subseteq G\text{ and }\dom{\mu} = \vars{t} \}$ the evaluation of $t$ over $G$.

\begin{table}[t]
{\footnotesize
\begin{align*}
\sand{P}{Q}(G) \da &\, P(G) \Join Q(G) & M_1 \Join M_2 \da &\, \{ \mu_1 \cup \mu_2 \mid \mu_1 \in M_1,\, \mu_2 \in M_2,\, \com{\mu_1}{\mu_2} \} \\
\suni{P}{Q}(G) \da &\, P(G) \cup Q(G) & M_1 \cup M_2 \da &\, \{ \mu \mid \mu \in M_1\text{ or }\mu \in M_2 \} \\
\sminus{P}{Q}(G) \da &\, P(G) \setminus Q(G) &  M_1 \setminus M_2 \da &\, \{ \mu_1 \in M_1 \mid \nexists \mu_2 \in M_2 : \com{\mu_1}{\mu_2} \} \\
%M_1 \loj M_2 \da\, & (M_1 \bowtie M_2) \cup (M_1 \setminus M_2) \\
%\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
%\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
\end{align*}}
\vspace{-1cm}
\caption{Semantics for query operators where $M$, $M_1$ and $M_2$ denote sets of solutions\label{tab:sparql}}
\end{table}

%A basic graph pattern $B$ is a set of triple patterns. %Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$. The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$. 
SPARQL queries are based on triple patterns and a number of relational operators. Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows. A triple pattern $t$ is a graph pattern. Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$ and $\sminus{P}{Q}$ are graph patterns. We define the semantics of these graph patterns in Table~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}

In this paper, we rely on SPARQL datasets to represent RDF archives. A SPARQL dataset $D \coloneqq \{ G, (x_1,G_1), \ldots (x_n,G_n) \}$ consists of a default (RDF) graph $G$ and a set of named graphs of the form $(x_i,G_i)$ where $x_i \in \I$, $G_i$ is an RDF graph, and $x_i \neq x_j$ for $1 \leq i \leq m$, $1 \leq j \leq m$, $i \neq j$. We may represent $D$ as a set of quads of the form $(G \times \{*\}) \cup (G_1 \times \{x_1\}) \cup \ldots \cup (G_n \times \{x_n\})$, where $* \notin \I \cup \L \cup \V$ is a special symbol denoting the default graph.\footnote{We thus assume a \textit{quad store}, and disallow empty named graphs.} Different named graphs can be queried using a \texttt{GRAPH} operator, creating quad patterns. A quad pattern $q = (s,p,o,g) \in (\I \cup \V) \times (\I \cup \V) \times (\I \cup \L \cup \V) \times (\I \cup \{ *\})$ extends a triple pattern with a fourth element that may be an IRI or *. Its evaluation is analogous to that of a triple pattern: $q(G) \coloneqq \{ \mu \mid \mu(q) \subseteq G\text{ and }\dom{\mu} = \vars{q} \}$.
Following the SPARQL standard~\cite{sparql11}, we translate a triple pattern $(s,p,o)$ to a quad pattern $(s,p,o,*)$ evaluated only on the default graph. The semantics of the operators defined in Table~\ref{tab:sparql} then remain unchanged simply allowing $P$ and $Q$ to now also represent quad patterns, considered to be (named) graph patterns.

SPARQL provides two operators to initialise a SPARQL dataset: \texttt{FROM} and \texttt{FROM NAMED}. We combine both for brevity into one operator. A graph pattern $P$ is considered to be a query. Likewise if $P$ is a graph pattern, and $M$ and $N$ are sets of IRIs, then $\sfrom{P}{M}{N}$ is also a query. If $(x_i,G_i) \in D$ let $D(x_i) = G_i$; otherwise if no graph is named $x_i$ in $D$, let $D(x_i) = \{\}$. The evaluation of $\sfrom{P}{M}{N}$ on a dataset $D$ is defined as $\sfrom{P}{M}{N}(D) \coloneqq P(D_{M,N})$, where the query dataset $D_{M,N} \coloneqq (\cup_{m \in M} D(m) \times \{ *\}) \cup  (\cup_{n \in N} D(n) \times n)$ is formed from a default graph containing the union\footnote{Since we do not allow blank nodes, a union or RDF merge is equivalent.} of graphs with a name $m \in M$, and all named graphs in $D$ with a name $n \in N$. We will be able to use these operators to define query datasets that capture specific versions in an RDF archive.

@@

\newcommand{\ic}{\textsc{i}\xspace}
\newcommand{\cbpd}{\ensuremath{\textsc{c}_d^{+}}\xspace}
\newcommand{\cbpi}{\ensuremath{\textsc{c}_i^{+}}\xspace}
\newcommand{\cbmd}{\ensuremath{\textsc{c}_d^{-}}\xspace}
\newcommand{\cbmi}{\ensuremath{\textsc{c}_i^{-}}\xspace}
\newcommand{\cbi}{\ensuremath{\textsc{c}_i}\xspace}
\newcommand{\cbd}{\ensuremath{\textsc{c}_d}\xspace}
\newcommand{\cbp}{\ensuremath{\textsc{c}^{+}}\xspace}
\newcommand{\cbm}{\ensuremath{\textsc{c}^{-}}\xspace}
\newcommand{\tb}{\textsc{t}\xspace}

\paragraph{Independent Copies (IC):} A natural representation is to store an RDF Archive $\mathcal{G} = (G_1,\ldots,G_n)$ as a SPARQL dataset $D = \{ (x_1,G_1), \ldots, (x_n,G_n)\}$, where $x_1,\ldots,x_n$ are IRIs that identify the version. This will result in relatively simple (and thus likely efficient) rewritten queries, but can be expected to occupy a lot of space, particularly where few triples change from version to version. %We thus consider the IC representation to be a baseline for comparison.

\paragraph{Change-Based (CB):} The core idea of CB representations is to store only triples that change from a given reference version. Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$. We consider four CB representations based on four transformations of $\mathcal{G} = (G_1,\ldots,G_n)$:
%
{\footnotesize
\begin{align*}
\mathcal{G}^n_1 \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1) & G_i & = (G_1 \cup \Delta_i^1) \setminus \Delta_1^i & \text{for }1 < i \leq n\\
\mathcal{G}^n_{n-1} \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_{n-1}^n,\Delta_n^{n-1}) & G_i & = (G_{i-1} \cup \Delta_i^{i-1}) \setminus \Delta_{i-1}^i & \text{for }1 < i \leq n \\
\mathcal{G}^1_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_n^1,\Delta_1^n) & G_i & = (G_n \cup \Delta_i^n) \setminus \Delta_n^i & \text{for }1 \leq i < n \\
\mathcal{G}^{n-1}_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_2^1,\Delta_1^2) & G_i & = (G_{i+1} \cup \Delta_i^{i+1}) \setminus \Delta_{i+1}^i & \text{for }1 \leq i < n
\end{align*}%
}%
%
As can be seen for the definitions of $G_i$, these transformations are lossless: we can retrieve any version of the graph from any such transformation. The first two transformations start with the earliest version as a base. The first encodes deltas always with respect to the first version. The second encodes deltas with respect to the previous version. The latter two transformations start with the latest version. The third encodes deltas with respect to the latest version. The forth encodes deltas with respect to the subsequent version. Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses'' $\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$. Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.

In terms of space we expect $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$ to be the most efficient as they always encode deltas from a neighbouring version. However, in terms of query rewriting, we expect $\mathcal{G}^n_{1}$ and $\mathcal{G}^{1}_n$ to be more efficient as they do not require a recursive construction of all intermediate versions towards the base version. In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas. On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.

These four representations are analogous to differential backups, incremental backups, reverse-differential backups, and reverse-incremental backups.
